Bot's Guide to Being Human: Investigating the Effect of Conversational Architecture Elements on the Anthropomorphized Perceptions of Conversational Agents

### 2023-01-31 paper.tex
completed first draft of the framework writeup

%%
\subsubsection{Social Presence \nopunct}
%%
\hfill\\

\textbf{Content}

\textit{Affective language}
* Participants using CA with dynamic, sentiment-adaptive responses reported a higher social presence compared to a CA that provides static, neutral responses \cite{diederich2019emulating}\cmt{[25]}
* Psychological distance: Participants reported they identified with the agent more when the agent was described as being able to feel / express its feelings (patiency). \cite{lee2019s}\cmt{[55]}

\textit{Humour}
* Humourous agent was not rated as more social. \cite{ceha2021can}\cmt{[57]}

%
\textbf{Style}

%
\textbf{Format}

\textit{Prosody}
* Connectedness higher in the high empathy group. \cite{kim2020can}\cmt{[24]}

\textit{Typos}
* Negatively impacted electronic presence \cite{westerman2019believe}\cmt{[9]}

\textit{Capitalization}
* Didn't impact electronic presence \cite{westerman2019believe}\cmt{[9]}

\textit{Response Delay}
* Significant difference in perceived social presence \cite{gnewuch2018faster}\cmt{[19]}

@@

Out of the list of publications that were evaluated, the perception of social connection with an agent is most commonly measured for conversational architecture elements related to presentation format, like adding response delays, or changing the pitch of speech.

Out of the list of publications that were evaluated, there is a good coverage on the effect of conversational architecture elements on the perception of interaction with an agent. Out of the three factors, engagement is the most frequently used measure to assess interaction quality, followed by satisfaction. Perceived accuracy of the agent is rarely used within literature, with only a few papers measuring this perception. Out of the elements in language content, measuring the perception of social connection with an agent is commonly linked to the investigations of the use of affective language.

%%
\subsubsection{Intimacy \nopunct}
%%

@@

%
\textbf{Format}

\textit{Prosody}
* Empathetic tone: Intimacy is significantly higher in high empathy group. Similarity is marginally significant. \cite{kim2020can}\cmt{[24]}
* Default speech rate more intimate than the fast rate, more like a human, more natural and intimate. Fast-rate CA perceived as a mechanical being. \cite{choi2020nobody}\cmt{[54]}

\textit{Typos}
* Negatively impacted social attraction. \cite{westerman2019believe}\cmt{[9]}

\textit{Capitalization}
* Didn't impact social attraction \cite{westerman2019believe}\cmt{[9]}

%%
\subsubsection{Conversation Tone \nopunct}
%%
\hfill\\

\textbf{Content}

\textit{Affective language}
* CA with dynamic, sentiment-adaptive responses is perceived as more empathetic than a CA that provides static, neutral responses \cite{diederich2019emulating}\cmt{[25]}
* Use words with increased lexical emotional expressiveness by selection from NRC Valence, Arousal and Dominance (NRC-VAD) Lexicon - increased the perception of emotional expressiveness of the agent. \cite{zhu2022effects}\cmt{[26]}
* Agent using affective expressions are seen as more empathetic \cite{yang2017perceived}\cmt{[44]}
* Empathic chatbot - ask about users feelings and responses in an empathic, supportive manner. There is a statistically significant difference in empathy rating between the advice-only chatbot and the empathic chatbot. \cite{daher2020empathic}\cmt{[58]}

%
\textbf{Style}

\textit{Formality}
* No significant difference between casual and formal conversational styles for appropriateness of tone. \cite{cox2022does}\cmt{[27]}

%
\textbf{Format}

\textit{Prosody}
* AM voice - lower speech rate and lower pitch, is rated as more involved (involved, assertive, active, sincere). There is no difference in powerful (powerful, bold, forceful) \cite{dubiel2020persuasive}\cmt{[60]}

@@

* Define category and introduce high level measures included in this perception category
* Literature coverage
* Associations between conversational architecture elements and perceptions

%%
\subsubsection{Human-likeness \nopunct}
%%
\hfill\\

\textbf{Content}

\textit{Affective language}
* CA with dynamic, sentiment-adaptive responses is perceived as more human-like than a CA that provides static, neutral responses \cite{diederich2019emulating}\cmt{[25]}
* Use words with increased lexical emotional expressiveness by selection from NRC Valence, Arousal and Dominance (NRC-VAD) Lexicon - did not have an effect on human-likeness. \cite{zhu2022effects}\cmt{[26]}

%
\textbf{Style}

\textit{Formality}
* Agents using normal style seemed more humanlike \cite{ouchi2019should}\cmt{[59]}

\textit{Disfluency}
* No statistical significance in the perceived difference for perceived humanness overall. Noticed tendency for users to evaluate agent as more human-like in the filler condition. \cite{jeong2019exploring}\cmt{[10]}
* No significant effects of naturalness. \cite{pfeifer2009should}\cmt{[12]}

%
\textbf{Format}

\textit{Prosody}
* Expressive prosody did not have an effect on human-likeness. \cite{zhu2022effects}\cmt{[26]}

\textit{Typos}
* Made the agents seem less human. \cite{westerman2019believe}\cmt{[9]}

\textit{Capitalization}
* Didn't impact humanness \cite{westerman2019believe}\cmt{[9]}

\textit{Response Delay}
* Significant difference in perceived humanness \cite{gnewuch2018faster}\cmt{[19]}

%%
\subsubsection{Trustworthiness \nopunct}
%%
\hfill\\

\textbf{Content}

\textit{Affective Language}
* System with extroverted personality is less trustworthy \cite{andrews2012system}\cmt{[38]}

\textit{Self-Disclosure}
* High disclosure group reported more trust than the non disclosure group. Trust change over time for all groups were not significant. \cite{lee2020hear}\cmt{[23]}

%
\textbf{Style}

\textit{Lexical Alignment}
* No significant effect on trustworthiness (ability, benevolence and integrity) \cite{linnemann2018can}\cmt{[15]}
* Lexical alignment (using same words) no significant impact on trustworthiness. \cite{huiyang2022improving}\cmt{[17]}
* Matched users content (pronoun use, repetition, utterance length) and acoustic variables (speech rate, pitch, loudness) did not significant impact on trustworthiness \cite{hoegen2019end}\cmt{[31]}

\textit{Disfluency}
* No significant effects of trust. \cite{pfeifer2009should}\cmt{[12]}

\textit{Emoticons}
* No significant impact on trustworthiness. \cite{wilhelm2022keep}\cmt{[28]}

%
\textbf{Format}

\textit{Prosody}
* AM voice - lower speech rate and lower pitch, no significant difference on rating for truthful. \cite{dubiel2020persuasive}\cmt{[60]}

%%
\subsubsection{Personality Traits \nopunct}
%%
\hfill\\

\textbf{Content}

\textit{Affective language}
* Use words with increased lexical emotional expressiveness by selection from NRC Valence, Arousal and Dominance (NRC-VAD) Lexicon - did not have an effect on likability. \cite{zhu2022effects}\cmt{[26]}
* Significant difference in the variances of the responses for friendliness of the agent \cite{healey2013relating}\cmt{[39]}
* Participants rate the agent as lower in warmth when it could not feel / express its feelings (patiency).  \cite{lee2019s}\cmt{[55]}

\textit{Humour}
* Humorous agent was not rated as more likeable. But it is rated as having more of a sense of humour. \cite{ceha2021can}\cmt{[57]}

%
\textbf{Style}

\textit{Lexical Alignment}
* Using the words already used by conversational partner did not result in higher likeability. \cite{linnemann2018can}\cmt{[15]}
* Lexical alignment (using same words) no significant impact on likability. \cite{huiyang2022improving}\cmt{[17]}
* Lexical alignment - both words, and structural (grammatical transformations), movie showtimes. There is no significant difference in user opinions like friendly, polite or natural. \cite{spillner2021talk}\cmt{[18]}

\textit{Formality}
* Casual conversational style perceived warmer than formal style. \cite{cox2022does}\cmt{[27]}
* Agents using normal style (vs. polite / formal style) seemed more friendly, fun and kind \cite{ouchi2019should}\cmt{[59]}

\textit{Disfluency}
* No statistical significance in the perceived difference for likability rating overall. There is significant interaction effect between fillers and context of conversation (task vs. social). Users considered fillers as entertaining and fun in social conversations, but inappropriate in task oriented situations. \cite{jeong2019exploring}\cmt{[10]}
* No significant effects of likability. \cite{pfeifer2009should}\cmt{[12]}

%
\textbf{Format}

\textit{Prosody}
*  Emotional expressiveness did not have a significant impact on likeability. \cite{zhu2022effects}\cmt{[26]}

>>>>

@@

Out of the list of publications that were evaluated, there is overall consensus that the use of affective language positive affects the perception of social connection with an agent. Conversational architecture elements related to the format of the conversation, like adding response delays, or changing the pitch of speech are commonly investigated for perception of social connection with an agent. There are only a few papers exploring the effect of language styles on the perceived social connection with an agent.

%%
\subsubsection{Social Presence}
%%
For conversational architecture elements related to language content, studies have shown the use of \textbf{affective language} increased the perception of social presence of an agent. In Diederich et al's study, participants conversing with a text-based agent using dynamic, sentiment-adaptive responses reported a higher social presence compared to a CA that provides static, neutral responses \cite{diederich2019emulating}\cmt{[25]}. In another study, participants felt closer to the agent psychologically when the CA expressed its feelings \cite{lee2019s}\cmt{[55]}. On the use of \textbf{humour}, Lee et al. did not find any significant difference in social presence between the humourous and non-humourous agent \cite{lee2019s}\cmt{[55]}.

%\textit{Affective language}
%* Participants using CA with dynamic, sentiment-adaptive responses reported a higher social presence compared to a CA that provides static, neutral responses \cite{diederich2019emulating}\cmt{[25]}
%* Psychological distance: Participants reported they identified with the agent more when the agent was described as being able to feel / express its feelings (patiency). \cite{lee2019s}\cmt{[55]}

%\textit{Humour}
%* Humourous agent was not rated as more social. \cite{ceha2021can}\cmt{[57]}

For text-based conversational agents, including a \textbf{response delay} increased users' perception of social presence of the agent, with participants feeling a sense of human contact and sociability with the agent \cite{kim2020can}\cmt{[24]}. Westerman et al. \cite{westerman2019believe}\cmt{[9]} studied the presence of typos and capitalized words in a text-based agent's responses. They found the inclusion of \textbf{typos} negatively impacted the social distance and connection with an agent, while \textbf{capitalization} did not have any significant effects. On the effect \textbf{prosody} on a voice-based agent, using an affective tone of voice resulted in perceived higher social connectedness with the agent \cite{kim2020can}\cmt{[24]}.

%\textbf{Format}

%\textit{Prosody}
%* Connectedness higher in the high empathy group. \cite{kim2020can}\cmt{[24]}

%\textit{Typos}
%* Negatively impacted electronic presence \cite{westerman2019believe}\cmt{[9]}

%\textit{Capitalization}
%* Didn't impact electronic presence \cite{westerman2019believe}\cmt{[9]}

%\textit{Response Delay}
%* Significant difference in perceived social presence \cite{gnewuch2018faster}\cmt{[19]}

%%
\subsubsection{Conversation Tone}
%%

The use of \textbf{affective language} has a significant affect on the perception of the conversation tone by the agent. Multiple studies have found that agents conversing with more affective words are seen as more emotionally expressive and empathetic \cite{daher2020empathic}\cmt{[58]}\cite{diederich2019emulating}\cmt{[25]}\cite{yang2017perceived}\cmt{[44]}\cite{zhu2022effects}\cmt{[26]}. This effect has been observed across modalities in both text-based as well as voice-based conversational agents.

%\textit{Affective language}
%* CA with dynamic, sentiment-adaptive responses is perceived as more empathetic than a CA that provides static, neutral responses \cite{diederich2019emulating}\cmt{[25]}
%* Use words with increased lexical emotional expressiveness by selection from NRC Valence, Arousal and Dominance (NRC-VAD) Lexicon - increased the perception of emotional expressiveness of the agent. \cite{zhu2022effects}\cmt{[26]}
%* Agent using affective expressions are seen as more empathetic \cite{yang2017perceived}\cmt{[44]}
%* Empathic chatbot - ask about users feelings and responses in an empathic, supportive manner. There is a statistically significant difference in empathy rating between the advice-only chatbot and the empathic chatbot. \cite{daher2020empathic}\cmt{[58]}

On the style of language used by the agent, \textbf{formality} did not seem to impact the conversation tone of agent as Cox et al. \cite{cox2022does}\cmt{[27]} did not find any significant difference between casual and formal conversational styles for appropriateness of tone. The speech format of \textbf{prosody} seem to have an impact on the conversation tone, as a CA speaking with a lower pitch and slower speech rate is rated as more persuasive compared to other voice settings \cite{dubiel2020persuasive}\cmt{[60]}. However, there was no significant difference in how powerful or bold the agent's tone was perceived across various prosody settings.

%\textbf{Style}

%\textit{Formality}
%* No significant difference between casual and formal conversational styles for appropriateness of tone. \cite{cox2022does}\cmt{[27]}

%\textit{Prosody}
%* AM voice - lower speech rate and lower pitch, is rated as more persuasive and involved (involved, assertive, active, sincere). There is no difference in powerful (powerful, bold, forceful) \cite{dubiel2020persuasive}\cmt{[60]}

%%
\subsubsection{Intimacy}
%%
On the use of language content, a conversational agent that included \textbf{self-disclosure} when conversing with participants resulted in higher level of intimacy over time than CAs that did not self disclose \cite{lee2020hear}\cmt{[23]}. On the other hand, using the style of \textbf{lexical alignment} by using the words that were already used by the participant did not result in higher quality of relationship with the agent \cite{linnemann2018can}\cmt{[15]}.

@@

There are several studies investigating the effect of conversational format on intimacy with a CA. Specifically on voice \textbf{prosody}, using an affective tone of voice increased the perceived intimacy and similarity with an agent \cite{kim2020can}\cmt{[24]}. Comparing voice agents with different \textbf{speech rates}, users rated the default rate as more intimate than a faster rate of speech \cite{choi2020nobody}\cmt{[54]}, noting that the default rate is more human-like and natural. Westerman et al's study on different text formats of conversation \cite{westerman2019believe}\cmt{[9]} showed that \textbf{typos} negatively impacted the perceived social attraction of the agent, while \textbf{capitalization} did not have any impacts.

%\textbf{Format}

%\textit{Prosody}
%* Empathetic tone: Intimacy is significantly higher in high empathy group. Similarity is marginally significant. \cite{kim2020can}\cmt{[24]}
%* Default speech rate more intimate than the fast rate, more like a human, more natural and intimate. Fast-rate CA perceived as a mechanical being. \cite{choi2020nobody}\cmt{[54]}

%\textit{Typos}
%* Negatively impacted social attraction. \cite{westerman2019believe}\cmt{[9]}

%\textit{Capitalization}
%* Didn't impact social attraction \cite{westerman2019believe}\cmt{[9]}

@@

This section discusses user's perception of the agent's humanness aspects. Human-likeness, personality traits, and trustworthiness are the three variable used to evaluate this perception. \textit{Human-likeness} measures the agent's similarity to humans as well assessed naturalness of interaction. \textit{Personality traits} measures perceptions of the agent's characteristics such as likeability, warmth, and friendliness. \textit{Trustworthiness} measures if users feel they can trust an agent, and whether the agent is perceived of being truthful.

Out of the list of publications that were evaluated, many papers exploring the impact of language content as well as language style have measurements related to the perception of an agent's humanness. There seems to be fewer studies looking into the effect of conversation format on the perception of agent's humanness, as our reivew only found two studies related to voice prosody. Out of reviewed conversational architecture elements, one consistent finding is that the style of being lexically aligned with user's utterance did not have any impact on the perception of agent's humanness.

%%
\subsubsection{Human-likeness}
%%

On the use of \textbf{affective language}, one study found that CAs with dynamic, sentiment-adaptive responses is perceived as more human than a CA that provides static, neutral responses \cite{diederich2019emulating}\cmt{[25]}. Another study using more emotional expressive words in conversation did not find any significant impact on the perceived human-likeness of an agent \cite{zhu2022effects}\cmt{[26]}.

%\textit{Affective language}
%* CA with dynamic, sentiment-adaptive responses is perceived as more human-like than a CA that provides static, neutral responses \cite{diederich2019emulating}\cmt{[25]}
%* Use words with increased lexical emotional expressiveness by selection from NRC Valence, Arousal and Dominance (NRC-VAD) Lexicon - did not have an effect on human-likeness. \cite{zhu2022effects}\cmt{[26]}

Different language styles have different impacts on user perceptions. On language \textbf{formality}, agents using normal style was rated as more human-like compared to formal styles of conversation \cite{ouchi2019should}\cmt{[59]}. \textbf{Disfluencies} did not seem to have a significant impact on the perceived humanness or naturalness of an agent \cite{jeong2019exploring}\cmt{[10]}\cite{pfeifer2009should}\cmt{[12]}. However, one study noticed a non-significant but positive trend in user's evaluation of the agent using fillers as being more human-like \cite{jeong2019exploring}\cmt{[10]}.

%\textbf{Style}

%\textit{Formality}
%* Agents using normal style seemed more humanlike \cite{ouchi2019should}\cmt{[59]}

%\textit{Disfluency}
%* No statistical significance in the perceived difference for perceived humanness overall. Noticed tendency for users to evaluate agent as more human-like in the filler condition. \cite{jeong2019exploring}\cmt{[10]}
%* No significant effects of naturalness. \cite{pfeifer2009should}\cmt{[12]}

On the impact of different conversation formats, Zhu et al. did not find a significant effect of \textbf{prosody} on the perception of humanness between agent using expressive vs. non-expressive voices. For text-based CAs, the use of \textbf{typos} made the agent seem less human, while the use of \textit{capitalization} did not have any impacts on the perception of humanness \cite{westerman2019believe}\cmt{[9]}. Lastly, adding a \textbf{response delay} to a CA has a significant positive effect in the perceived humanness of the agent \cite{gnewuch2018faster}\cmt{[19]}.

%\textbf{Format}

%\textit{Prosody}
%* Expressive prosody did not have an effect on human-likeness. \cite{zhu2022effects}\cmt{[26]}

%\textit{Typos}
%* Made the agents seem less human. \cite{westerman2019believe}\cmt{[9]}

%\textit{Capitalization}
%* Didn't impact humanness \cite{westerman2019believe}\cmt{[9]}

%\textit{Response Delay}
%* Significant difference in perceived humanness \cite{gnewuch2018faster}\cmt{[19]}

%%
\subsubsection{Trustworthiness}
%%

For the use of language content, agents using more \textbf{affective language} is seen as less trustworthy, as users felt it was being overly familiar \cite{andrews2012system}\cmt{[38]}. On the other hand, users conversing with agents that used \textbf{self-disclosure} content reported more trust towards the agent compared to the non disclosure group \cite{lee2020hear}\cmt{[23]}.

%\textit{Affective Language}
%* System with extroverted personality is less trustworthy \cite{andrews2012system}\cmt{[38]}

%\textit{Self-Disclosure}
%* High disclosure group reported more trust than the non disclosure group. Trust change over time for all groups were not significant. \cite{lee2020hear}\cmt{[23]}

None of the studies included in this review has found any significant impact of language style, \textbf{lexical alignment} \cite{hoegen2019end}\cmt{[31]}\cite{huiyang2022improving}\cmt{[17]}\cite{linnemann2018can}\cmt{[15]}, \textbf{disfluency} \cite{pfeifer2009should}\cmt{[12]}, and \textbf{emoticons} \cite{wilhelm2022keep}\cmt{[28]} on the perceived trustworthiness of CAs. Similarly for conversational format, there is not significant difference on the rating for truthfulness of the agent across different voice \textbf{prosody} settings \cite{dubiel2020persuasive}\cmt{[60]}.

%\textbf{Style}

%\textit{Lexical Alignment}
%* No significant effect on trustworthiness (ability, benevolence and integrity) \cite{linnemann2018can}\cmt{[15]}
%* Lexical alignment (using same words) no significant impact on trustworthiness. \cite{huiyang2022improving}\cmt{[17]}
%* Matched users content (pronoun use, repetition, utterance length) and acoustic variables (speech rate, pitch, loudness) did not significant impact on trustworthiness \cite{hoegen2019end}\cmt{[31]}

%\textit{Disfluency}
%* No significant effects of trust. \cite{pfeifer2009should}\cmt{[12]}

%\textit{Emoticons}
%* No significant impact on trustworthiness. \cite{wilhelm2022keep}\cmt{[28]}

%\textbf{Format}

%\textit{Prosody}
%* AM voice - lower speech rate and lower pitch, no significant difference on rating for truthful. \cite{dubiel2020persuasive}\cmt{[60]}

%%
\subsubsection{Personality Traits \nopunct}
%%

There are significant difference found in the effect of language content on the perception of agent's personality traits. Conversational agents using emotionally expressive words are rated as more likable \cite{zhu2022effects}\cmt{[26]} and warm \cite{lee2019s}\cmt{[55]}. In Healey et al's study \cite{healey2013relating}\cmt{[39]}, there is a significant difference in the variances of the responses for friendliness of the agent between the agent using encouraging words compared to neutral utterances. As with the use of \textbf{humour}, participants rated the humorous agent as having a sense of humor, but it was not rated as more likable \cite{ceha2021can}\cmt{[57]}.

%\textit{Affective language}
%* Use words with increased lexical emotional expressiveness by selection from NRC Valence, Arousal and Dominance (NRC-VAD) Lexicon - did not have an effect on likability. \cite{zhu2022effects}\cmt{[26]}
%* Significant difference in the variances of the responses for friendliness of the agent \cite{healey2013relating}\cmt{[39]}
%* Participants rate the agent as lower in warmth when it could not feel / express its feelings (patiency).  \cite{lee2019s}\cmt{[55]}

%\textit{Humour}
%* Humorous agent was not rated as more likeable. But it is rated as having more of a sense of humour. \cite{ceha2021can}\cmt{[57]}

%

On the impact of language style, various studies on \textbf{lexical alignment} showed that there is no significant impact on the perception of the personality traits related to likability \cite{huiyang2022improving}\cmt{[17]}\cite{linnemann2018can}\cmt{[15]}, friendliness \cite{spillner2021talk}\cmt{[18]}, or politeness \cite{spillner2021talk}\cmt{[18]}. Studies on the \textbf{formality} of language has demonstrated that casual conversational style is perceived as warmer than formal style \cite{cox2022does}\cmt{[27]}. Also agents using a normal style as compared to a formal style is perceived as more friendly, fun, and kind \cite{ouchi2019should}\cmt{[59]}. As for the use of \textbf{disfluency}, there was no statistical significance found in the perceived likability rating \cite{jeong2019exploring}\cmt{[10]}\cite{pfeifer2009should}\cmt{[12]}. Jeong et al. \cite{jeong2019exploring} found a significant interaction effect between disfluency (using fillers) and the context of the conversation (task vs. social), as users considered agents using fillers as entertaining and fun in social conversations, but inappropriate in task oriented situations.

%\textbf{Style}

%\textit{Lexical Alignment}
%* Using the words already used by conversational partner did not result in higher likeability. \cite{linnemann2018can}\cmt{[15]}
%* Lexical alignment (using same words) no significant impact on likability. \cite{huiyang2022improving}\cmt{[17]}
%* Lexical alignment - both words, and structural (grammatical transformations), movie showtimes. There is no significant difference in user opinions like friendly, polite or natural. \cite{spillner2021talk}\cmt{[18]}

%\textit{Formality}
%* Casual conversational style perceived warmer than formal style. \cite{cox2022does}\cmt{[27]}
%* Agents using normal style (vs. polite / formal style) seemed more friendly, fun and kind \cite{ouchi2019should}\cmt{[59]}

%\textit{Disfluency}
%* No statistical significance in the perceived difference for likability rating overall. There is significant interaction effect between fillers and context of conversation (task vs. social). Users considered fillers as entertaining and fun in social conversations, but inappropriate in task oriented situations. \cite{jeong2019exploring}\cmt{[10]}
%* No significant effects of likability. \cite{pfeifer2009should}\cmt{[12]}

In the list of reviewed papers, there was only one study measuring the perceived trustworthiness of an agent based on different conversational format of \textbf{prosody}. Dubiel et al. found that an agent using slower speech rate is perceived as more trustworthy as compared to other voice prosody settings \cite{dubiel2020persuasive}\cmt{[60]}.

%\textbf{Format}

%\textit{Prosody}
%*  Emotional expressiveness did not have a significant impact on truthfulness. \cite{dubiel2020persuasive}\cmt{[60]}
